---
title: "Bulk Met Driver Prep for Multi-Site Workflows"
---

```{r}
#| include: false
knitr::opts_chunk$set(eval = FALSE)
```

::: {.callout-note}
## What you will learn

- How to prepare CF-standard met driver files for hundreds of sites
  in a single pass using `cae_build_met_drivers()`
- Why the bulk approach is dramatically more efficient than looping
  over `cae_fetch_point()`
- How to set up a sites table and configure the output for use with
  PEcAn or standalone ecosystem model runs
:::

## Background

Ecosystem modeling projects often need met drivers at many sites --
the CCMMF project, for example, runs SIPNET at 198 design points
across California. Preparing met files one site at a time with
`cae_fetch_point()` is straightforward but wasteful: each call reads
the entire WRF grid from S3 independently, so 198 sites x 8 variables
x 20 years means 31,680 S3 requests.

`cae_build_met_drivers()` solves this by calling `cae_fetch_points()`
(plural) under the hood. The full WRF grid is read from S3 once per
variable per year, and all sites are extracted from that grid in a
single vectorized operation. The same 198-site, 20-year job now takes
160 S3 requests.

## Preparing the sites table

The input is a data.frame with three required columns: `site_id`,
`lon`, and `lat`. Any additional columns are ignored.

```{r}
# read CCMMF design points
sites <- read.csv("design_points.csv")
head(sites)
#>   UniqueID          site_id      lat       lon          pft
#> 1  1301797 8773c58306e8d9a0 32.71585 -115.4716 annual crop
#> 2  1300446 f2ce12d75f3b2cc6 32.75557 -115.4159 annual crop
#> ...

# cae_build_met_drivers needs site_id, lon, lat
# the site_id column is already present in this table
```

If your sites come from PEcAn's BETY database or another source, just
make sure you have those three columns:

```{r}
# manual setup for a few sites
sites <- data.frame(
  site_id = c("fresno", "la", "sacramento"),
  lon = c(-119.77, -118.24, -121.49),
  lat = c(36.75, 34.05, 38.58)
)
```

## Building the drivers

```{r}
library(caladaptaer)

result <- cae_build_met_drivers(
  sites     = sites,
  model     = "CESM2",
  scenario  = "ssp370",
  start_year = 2050,
  end_year   = 2055,
  outdir    = "met_drivers"
)
```

Progress output looks like:

```
Year 2050 (1 of 6)
  fetching t2
Reading grid (1 S3 fetch for all 3 sites)
Extracting 3 sites
  fetching psfc
  ...
  wrote met_drivers/fresno/CESM2.ssp370.2050.nc
  wrote met_drivers/la/CESM2.ssp370.2050.nc
  wrote met_drivers/sacramento/CESM2.ssp370.2050.nc
Year 2051 (2 of 6)
  ...
```

The result is a data.frame tracking what was written:

```{r}
head(result)
#>    site_id year                                        file
#> 1   fresno 2050   met_drivers/fresno/CESM2.ssp370.2050.nc
#> 2       la 2050       met_drivers/la/CESM2.ssp370.2050.nc
#> 3 sacramento 2050 met_drivers/sacramento/CESM2.ssp370.2050.nc
```

## Output file structure

Each NetCDF file contains 9 CF-standard variables at hourly resolution
for one site and one year:

| CF variable | Units | Source |
|-------------|-------|--------|
| `air_temperature` | K | t2 (direct) |
| `air_pressure` | Pa | psfc (direct) |
| `precipitation_flux` | kg m-2 s-1 | prec / 3600 |
| `specific_humidity` | 1 | q2 / (1 + q2) |
| `surface_downwelling_shortwave_flux_in_air` | W m-2 | swdnb (direct) |
| `surface_downwelling_longwave_flux_in_air` | W m-2 | lwdnb (direct) |
| `eastward_wind` | m s-1 | u10 (direct) |
| `northward_wind` | m s-1 | v10 (direct) |
| `wind_speed` | m s-1 | sqrt(u10^2 + v10^2) |

Dimensions are `latitude(1)`, `longitude(1)`, and `time(unlimited)`.
The time coordinate uses `seconds since {year}-01-01 00:00:00`.

## Skipping existing files

By default, `cae_build_met_drivers()` skips sites that already have
output files for a given year. If all sites in a year are complete, the
S3 fetch for that year is skipped entirely. Pass `overwrite = TRUE` to
regenerate everything.

```{r}
# re-run after interruption -- only missing files are written
result <- cae_build_met_drivers(
  sites, model = "CESM2", scenario = "ssp370",
  start_year = 2050, end_year = 2055,
  outdir = "met_drivers"
)
```

## Verifying the output

Spot-check a file to make sure values are physically reasonable:

```{r}
library(ncdf4)

nc <- nc_open("met_drivers/fresno/CESM2.ssp370.2050.nc")
t2 <- ncvar_get(nc, "air_temperature")
pr <- ncvar_get(nc, "precipitation_flux")
nc_close(nc)

# temperature: 200--350 K
range(t2)

# precip flux: 0--0.01 kg/m2/s (0--36 mm/hr, reasonable max for hourly)
range(pr)
```

## Efficiency comparison

For 198 sites x 8 variables x 20 years:

| Approach | S3 requests | Description |
|----------|:-----------:|-------------|
| `cae_fetch_point()` loop | 31,680 | one grid read per site per var per year |
| `cae_build_met_drivers()` | 160 | one grid read per var per year |
| PEcAn `download.CalAdaptWRF` | 160 | same count, uses tempdir() cache |

All three produce identical output. The choice depends on whether you
are working inside PEcAn (`download.CalAdaptWRF`) or outside it
(`cae_build_met_drivers`).

## Multiple models and scenarios

To build drivers for multiple GCMs or scenarios, loop over the
combinations:

```{r}
models <- c("CESM2", "CNRM-ESM2-1", "EC-Earth3-Veg")

for (m in models) {
  cae_build_met_drivers(
    sites, model = m, scenario = "ssp370",
    start_year = 2050, end_year = 2069,
    outdir = file.path("met_drivers", m)
  )
}
```

CESM2 supports SSP2-4.5 and SSP5-8.5 in addition to SSP3-7.0. All
other GCMs have SSP3-7.0 only.

## References

- Rahimi, S. et al. (2024). An overview of the Western United States
  Dynamically Downscaled Dataset (WUS-D3). *Geoscientific Model Development*,
  17, 2265--2286.
  [doi:10.5194/gmd-17-2265-2024](https://doi.org/10.5194/gmd-17-2265-2024)
- Dietze, M. C. et al. (2014). A quantitative assessment of a terrestrial
  biosphere model's data needs across North American biomes.
