---
title: "Accessing Cal-Adapt Data in R"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This is the full tour. Every exported function in caladaptaer, with a real
call and the output it gives back, all framed around a mid century window
(July 2050, SSP3-7.0) so the examples hang together. If you just want to get
a time series on screen in a couple of calls, start with
[Getting Started](getting_started.qmd) and come back here when you want the
whole surface.

The functions fall into three groups: **discovery** (what is on S3 and where),
**fetching** (pull grids and points into R), and **drivers** (turn WRF output
into CF standard NetCDF for ecosystem models). Argument notes here are short
on purpose. The full signature for every function, with defaults and return
types, lives in the [package reference](../reference/index.html).

## What is on the Analytics Engine

The [Cal-Adapt Analytics Engine](https://analytics.cal-adapt.org/) hosts about
1 PB of downscaled CMIP6 climate projections for California and the western US,
produced for the state's Fifth Climate Change Assessment. The data sits in
cloud optimized [Zarr](https://zarr.dev/) stores on AWS S3, public, no
credentials. caladaptaer reads those stores straight into R through GDAL's
`/vsis3/` virtual filesystem, so there is no Python, no API key, and nothing to
authenticate.

Two downscaling products are on offer:

**WRF CMIP6 (WUS-D3)**, dynamically downscaled by UCLA (Rahimi et al. 2024).
The Weather Research and Forecasting model runs hourly at three nested grids:
45 km (d01, western US), 9 km (d02, California and surroundings), and 3 km (d03,
California). Eight GCMs are downscaled, mostly under SSP3-7.0, with CESM2 also
carrying SSP2-4.5 and SSP5-8.5. There is also an ERA5 reanalysis driven run.
WRF gives you the full surface meteorology: temperature, pressure,
precipitation, humidity, radiation, and wind.

**LOCA2-Hybrid**, statistically downscaled by Scripps (Pierce et al.). Daily
output at 3 km for 15 GCMs across three SSPs, covering 1950 to 2100.

## Setup

```{r}
library(caladaptaer)
library(stars)
library(sf)
library(maps)
```

The Zarr driver needs GDAL 3.4 or newer:

```{r}
sf::sf_extSoftVersion()["GDAL"]
```

## Discovery

### The catalog

`cae_catalog()` pulls the CADCAT index, one row per variable per
model/scenario/resolution combo, and caches it for the session. Pass
`refresh = TRUE` to force a fresh download.

```{r}
cat <- cae_catalog()
nrow(cat)
names(cat)
```

The columns follow CMIP6 naming. This is the vocabulary every other function
speaks:

| Column | Meaning | Examples |
|--------|---------|----------|
| `activity_id` | Downscaling method | WRF, LOCA2 |
| `source_id` | Parent GCM | CESM2, MIROC6, EC-Earth3 |
| `experiment_id` | Scenario | ssp370, ssp245, historical |
| `table_id` | Temporal resolution | 1hr, day, mon |
| `variable_id` | Variable short name | t2, prec, pr, tasmax |
| `grid_label` | Spatial grid | d01 (45 km), d02 (9 km), d03 (3 km) |
| `member_id` | Ensemble member | r1i1p1f1 |
| `path` | S3 Zarr store | s3://cadcat/wrf/ucla/cesm2/... |

### Models, scenarios, variables

Three thin helpers read straight off the catalog so you do not have to filter
it by hand. Each takes an `activity` of `"WRF"` or `"LOCA2"` (or `NULL` for
both).

```{r}
cae_models(activity = "WRF")
```

```{r}
cae_models(activity = "LOCA2")
```

`cae_scenarios()` also takes a `model` to narrow it down. CESM2 carries the
widest set on the WRF side:

```{r}
cae_scenarios(activity = "WRF", model = "CESM2")
```

`cae_variables()` takes a `timescale` (`"1hr"`, `"day"`, `"mon"`). The hourly
WRF list is long because it includes the full radiation and hydrology fields:

```{r}
cae_variables(activity = "WRF", timescale = "1hr")
```

### Searching

`cae_search()` filters the catalog on any combination of `activity`, `model`,
`scenario`, `variable`, `timescale`, and `resolution`. Everything you pass is
ANDed together.

```{r}
cae_search(activity = "WRF", model = "CESM2",
           variable = "t2", timescale = "1hr", resolution = "d01")
```

### Checking a driver set before you fetch

Before you commit to building met drivers for a model and scenario, ask
whether the variables you need are actually there. `cae_check_variables()`
takes a `model` and `scenario` and reports, for the eight fields SIPNET wants,
whether each is directly available or derivable.

```{r}
cae_check_variables("CESM2", "ssp370")
```

`derivable` flags variables built from others, like total precipitation summed
from its convective and non convective parts. A clean run of `TRUE` in the
`available` column means you are good to go.

## Fetching

### A grid

`cae_fetch()` reads a Zarr store from S3 and hands back a `stars` object. Keep
the first read small by windowing with `start_time` and `end_time` (ISO strings
or POSIXct); a one hour window is one timestep. `resolution` picks the grid and
`timescale` picks `1hr` (WRF) or `day` (LOCA2).

```{r}
t2 <- cae_fetch("t2", model = "CESM2", scenario = "ssp370",
                start_time = "2050-07-01T00:00:00", end_time = "2050-07-01T00:00:00")
t2
```

Three dimensions come back: x and y in the WRF Lambert Conformal grid, and
time in UTC. Values are 2 meter air temperature in Kelvin.

### Mapping it

The 45 km domain covers most of western North America. Drop state lines on top
to orient the map.

```{r}
#| fig-width: 7
#| fig-height: 7
#| fig-alt: "2 meter air temperature over the western US at 45 km on 1 July 2050 under CESM2 SSP3-7.0, with cool Sierra and Cascade ridges, warm Southwest deserts, and a cooler Pacific coast. California outlined in black."
western <- st_as_sf(
  map("state", c("california", "oregon", "washington", "nevada",
                 "arizona", "utah", "idaho", "montana", "wyoming",
                 "colorado", "new mexico"), plot = FALSE, fill = TRUE)
)
ca <- st_as_sf(map("state", "california", plot = FALSE, fill = TRUE))
western_wrf <- st_transform(western, st_crs(t2))
ca_wrf <- st_transform(ca, st_crs(t2))

t2_c <- t2
t2_c[[1]] <- units::drop_units(t2[[1]]) - 273.15

plot(t2_c, main = "", axes = TRUE, reset = FALSE,
     col = hcl.colors(100, "YlOrRd", rev = TRUE),
     key.pos = 4, key.width = lcm(1.5), key.length = 0.8)
plot(st_geometry(western_wrf), add = TRUE, border = "grey30", lwd = 0.6)
plot(st_geometry(ca_wrf), add = TRUE, border = "black", lwd = 1.5)
```

For California only work at 9 or 3 km, pass `resolution = "d02"` or `"d03"`,
but mind the memory: one 3 km timestep is about 150 MB against 0.3 MB at
45 km.

### A point

`cae_fetch_point()` reads the grid and pulls the time series at the nearest
cell to a `lon`/`lat`. The transform from WGS84 to the WRF Lambert grid is
handled for you. Here is a summer day at Fresno:

```{r}
ts <- cae_fetch_point("t2", model = "CESM2", scenario = "ssp370",
                      lon = -119.77, lat = 36.75,
                      start_time = "2050-07-01T00:00:00",
                      end_time   = "2050-07-01T23:00:00")
head(ts)
```

You get a plain `data.frame` of `time` and `value`, so base plotting works
right away.

```{r}
#| fig-width: 8
#| fig-height: 4.5
#| fig-alt: "Hourly 2 meter air temperature at Fresno for 1 July 2050 under CESM2 SSP3-7.0, swinging from roughly 12 to 31 degrees C across the day."
ts$temp_c <- ts$value - 273.15
plot(ts$time, ts$temp_c, type = "l", lwd = 2, col = "#D62728",
     xlab = "time", ylab = "temperature (C)")
```

### Many points in one read

`cae_fetch_points()` is the design choice worth knowing. It reads the grid from
S3 once and extracts every site locally in a single pass. Three sites or ten
thousand, it is one S3 read per variable, not one read per site. Pass a
`data.frame` with `lon` and `lat` columns (and an optional `site_id`).

```{r}
#| message: true
sites <- data.frame(
  site_id = c("davis", "fresno", "sacramento"),
  lon = c(-121.74, -119.77, -121.49),
  lat = c(38.54, 36.75, 38.58)
)
mp <- cae_fetch_points("t2", model = "CESM2", scenario = "ssp370",
                       points = sites,
                       start_time = "2050-07-01T00:00:00",
                       end_time   = "2050-07-01T05:00:00")
head(mp, 8)
```

### Comparing models

Because every GCM rides the same downscaling framework, lining them up at one
location shows the inter model spread, a read on structural uncertainty in the
projections.

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Hourly 2 meter air temperature at Fresno for 1 July 2050 from every WRF GCM under SSP3-7.0. The curves share the cool night, warm afternoon shape but separate by a few degrees C in absolute level."
wrf_370 <- Filter(function(m) "ssp370" %in% cae_scenarios(activity = "WRF", model = m),
                  setdiff(cae_models(activity = "WRF"), "ERA5"))

model_ts <- lapply(wrf_370, function(m) {
  d <- cae_fetch_point("t2", model = m, scenario = "ssp370",
                       lon = -119.77, lat = 36.75,
                       start_time = "2050-07-01T00:00:00",
                       end_time   = "2050-07-01T23:00:00")
  d$model  <- m
  d$temp_c <- d$value - 273.15
  d
})
model_ts <- do.call(rbind, model_ts)

cols <- hcl.colors(length(wrf_370), "Dark 3")
plot(model_ts$time, model_ts$temp_c, type = "n", xlab = "time", ylab = "temperature (C)")
for (i in seq_along(wrf_370)) {
  d <- model_ts[model_ts$model == wrf_370[i], ]
  lines(d$time, d$temp_c, col = cols[i], lwd = 2)
}
legend("topright", legend = wrf_370, col = cols, lwd = 2, cex = 0.7, bty = "n")
```

### The full surface meteorology

SIPNET and most ecosystem models want several fields at once. The hourly WRF
data carries all of them. Here are the seven SIPNET inputs at Fresno, each one
its own `cae_fetch_point()` call:

| WRF name | Field | Units |
|----------|-------|-------|
| `t2` | 2 meter air temperature | K |
| `psfc` | Surface pressure | Pa |
| `prec` | Precipitation | mm/timestep |
| `q2` | Water vapor mixing ratio | kg/kg |
| `swdnb` | Downward shortwave radiation | W/m2 |
| `u10` | 10 meter eastward wind | m/s |
| `v10` | 10 meter northward wind | m/s |

```{r}
vars <- c("t2", "psfc", "prec", "q2", "swdnb", "u10", "v10")
met <- lapply(vars, function(v) {
  cae_fetch_point(v, model = "CESM2", scenario = "ssp370",
                  lon = -119.77, lat = 36.75,
                  start_time = "2050-07-01T00:00:00",
                  end_time   = "2050-07-01T23:00:00")
})
names(met) <- vars
sapply(met, function(d) round(mean(d$value), 2))
```

## Drivers

The last two functions turn fetched WRF output into CF standard NetCDF that
ecosystem models read directly. These write files to disk, so they are shown
here without running.

### One site, hand assembled

`cae_write_cf_netcdf()` takes a named list of `stars` objects (names are WRF
variable IDs), does the unit work (precipitation to flux, mixing ratio to
specific humidity, U and V to wind speed), and writes one CF-1.8 file.
`dt_seconds` sets the timestep length, default 3600 for hourly.

```{r}
#| eval: false
var_list <- list(
  t2    = cae_fetch("t2",    model = "CESM2", scenario = "ssp370",
                    start_time = "2050-01-01", end_time = "2050-12-31"),
  psfc  = cae_fetch("psfc",  model = "CESM2", scenario = "ssp370",
                    start_time = "2050-01-01", end_time = "2050-12-31"),
  prec  = cae_fetch("prec",  model = "CESM2", scenario = "ssp370",
                    start_time = "2050-01-01", end_time = "2050-12-31")
  # ... and the rest of the SIPNET set
)
cae_write_cf_netcdf(var_list, outfile = "fresno_2050.nc", dt_seconds = 3600)
```

### Many sites, end to end

`cae_build_met_drivers()` is the production path. Hand it a `data.frame` of
sites (`site_id`, `lon`, `lat`), a `model`, `scenario`, and a year range, and
it fetches every required variable once, extracts all sites, converts units,
and writes one CF NetCDF per site per year under `outdir`. `resolution`
defaults to `d01`; `overwrite` and `verbose` do what they say.

```{r}
#| eval: false
sites <- data.frame(
  site_id = c("site_0001", "site_0002", "site_0003"),
  lon = c(-121.74, -119.77, -121.49),
  lat = c(38.54, 36.75, 38.58)
)
cae_build_met_drivers(sites,
                      model = "CESM2", scenario = "ssp370",
                      start_year = 2050, end_year = 2050,
                      outdir = "met/", resolution = "d01")
```

This is the engine behind the production runs. For the full driver pipeline,
walked step by step, see [SIPNET Met Drivers](sipnet_met_drivers.qmd), and for
scaling it across many sites see [Bulk Met Drivers](bulk_met_drivers.qmd).

## Where to go next

- [SIPNET Met Drivers](sipnet_met_drivers.qmd) -- raw WRF to CF standard NetCDF
- [Bulk Met Drivers](bulk_met_drivers.qmd) -- the multi site driver path
- [Package reference](../reference/index.html) -- every function, every
  argument, every default

## References

- Rahimi, S. et al. (2024). An overview of the Western United States
  Dynamically Downscaled Dataset (WUS-D3). *Geoscientific Model Development*,
  17, 2265--2286.
  [doi:10.5194/gmd-17-2265-2024](https://doi.org/10.5194/gmd-17-2265-2024)
- Pierce, D. W. et al. Statistical downscaling using Localized Constructed
  Analogs (LOCA). *Journal of Hydrometeorology*, 15(6), 2558--2585.
- Cal-Adapt Analytics Engine: <https://analytics.cal-adapt.org/>
- CADCAT on AWS: <https://registry.opendata.aws/caladapt-coproduced-climate-data/>
