---
title: "Accessing Cal-Adapt Data in R"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Purpose

This guide describes the main data access functions in `caladaptaer`.
For the shortest path from installation to a time series, see
[Getting Started](getting_started.qmd).

These functions fall into three groups:

- **discovery** (what data are available),
- **fetching** (reading grids and point time series into R), and
- **coverage checks** (confirming variable availability before larger
  workflows).

All functions are documented, with signatures, defaults, and return types in
the [package reference](../reference/index.html) or in R help
(for example, `?cae_catalog`).

## What is on the Analytics Engine

The [Cal-Adapt Analytics Engine](https://analytics.cal-adapt.org/) hosts about
1 PB of downscaled CMIP6 climate projections for California and the western US,
produced for the state's Fifth Climate Change Assessment. The data are stored
as cloud-optimized [Zarr](https://zarr.dev/) arrays on AWS S3.

Zarr stores are chunked: a reader can request only the time and spatial chunks
needed for an analysis instead of downloading a full dataset. `caladaptaer`
uses GDAL's `/vsis3/` virtual filesystem to read those public S3 stores and
returns the requested subset as a native R object. Gridded reads return
`stars` objects; point extraction helpers return `data.frame`s.

This package provides native R access to the CAE stores.
It does not require Python, a CAE API key, or CAE authentication.

Access to two downscaling products is currently implemented:

**WRF CMIP6 (WUS-D3)** is dynamically downscaled using the
Weather Research and Forecasting (WRF) model (Rahimi et al. 2024).
The WRF downscaled runs are provided at hourly time steps on three nested grids:
45 km (d01, western US), 9 km (d02, California and surroundings), and 3 km (d03,
California). The WRF CMIP6 time domain is 1980--2100: historical GCM runs cover
1980--2014, and future scenario runs cover 2015--2100. Eight General
Circulation Models (GCMs) are downscaled under the Shared Socioeconomic Pathway
SSP3-7.0 scenario. CESM2 is also available under SSP2-4.5 and SSP5-8.5.

The Analytics Engine also includes an ERA5-driven WRF historical reconstruction.
This is a reanalysis-based product, not a future GCM scenario. The underlying
WRF-ERA5 evaluation covers 1980--2020 (Rahimi et al. 2022); in the Analytics
Engine, reanalysis products are described separately from GCM historical runs.

The downscaled WRF output includes a range of surface meteorology variables. These
vary by model, but all include temperature, pressure, precipitation, humidity,
solar radiation, and wind.

**LOCA2-Hybrid**, statistically downscaled by Scripps (Pierce et al.). Daily
output at 3 km for 15 GCMs across three SSPs, covering 1950 to 2100.

### Choosing between products

The products differ in ways that matter for downstream analysis:

| Product | Time step | Time domain | Ensemble | Use when |
|---|---|---|---|---|
| WRF CMIP6 | Hourly | 1980--2014 historical GCM runs; 2015--2100 SSP runs | Smaller GCM ensemble; SSP3-7.0 is broadest, with additional CESM2 scenarios | Analyses requiring sub-daily meteorology, pressure, radiation, wind, or historical/future simulations from the same downscaling framework. |
| WRF-ERA5 historical reconstruction | Hourly | 1980--2020 | Single reanalysis-driven run | Historical analyses requiring an observation-constrained reconstruction rather than a future climate scenario. |
| LOCA2-Hybrid | Daily | 1950--2014 historical GCM runs; 2015--2100 SSP runs | Broader GCM and SSP coverage | Analyses requiring a larger daily ensemble or daily temperature and precipitation extremes, without hourly driver variables. |

WRF CMIP6 provides historical and future simulations at the same spatial
resolution and with the same dynamical downscaling method, but the historical
GCM runs are still climate model simulations, not observations. The ERA5-driven
WRF reconstruction is historical only and is better suited when a reanalysis
baseline is needed.

### Product-comparison example

WRF and LOCA2-Hybrid differ by downscaling method, grid, timestep, and variable
names. The following non-evaluated example shows the package interface for
comparing hourly WRF `t2` with daily LOCA2-Hybrid `tasmax` and `tasmin` for the
same model, scenario, location, and week.

```{r}
#| eval: false
wrf <- cae_fetch_point(
    variable = "t2",
    model = "EC-Earth3",
    scenario = "ssp370",
    lon = -119.77,
    lat = 36.75,
    start_time = "2050-07-01T00:00:00",
    end_time = "2050-07-07T23:00:00"
)

tasmax <- cae_fetch_point(
    variable = "tasmax",
    model = "EC-Earth3",
    scenario = "ssp370",
    lon = -119.77,
    lat = 36.75,
    timescale = "day",
    resolution = "d03",
    start_time = "2050-07-01",
    end_time = "2050-07-07"
)

tasmin <- cae_fetch_point(
    variable = "tasmin",
    model = "EC-Earth3",
    scenario = "ssp370",
    lon = -119.77,
    lat = 36.75,
    timescale = "day",
    resolution = "d03",
    start_time = "2050-07-01",
    end_time = "2050-07-07"
)
```

The returned objects use the same `data.frame` structure, but the values should
not be treated as interchangeable. WRF `t2` is hourly near-surface air
temperature from a dynamical simulation; LOCA2-Hybrid `tasmax` and `tasmin` are
daily maximum and minimum temperatures from a statistical downscaling product.

## Setup

See [Getting Started](getting_started.qmd) for installation instructions. This
tutorial assumes `caladaptaer` and the supporting spatial packages are already
installed.

```{r}
library(caladaptaer)
library(stars)
library(sf)
library(maps)
```

The Zarr driver requires GDAL 3.4 or newer:

```{r}
sf::sf_extSoftVersion()["GDAL"]
```

## Discovery

### CAE catalog inventory

`cae_catalog()` returns the CADCAT index, one row per
model/scenario/variable/resolution combination, and caches it for the session.

```{r}
cat <- cae_catalog()
nrow(cat)
names(cat)
head(cat)
```

The columns of the catalog follow CMIP6 naming. All functions use this vocabulary:

| Column | Meaning | Examples |
|--------|---------|----------|
| `activity_id` | Downscaling method | WRF, LOCA2 |
| `source_id` | Parent GCM | CESM2, MIROC6, EC-Earth3 |
| `experiment_id` | Scenario | ssp370, ssp245, historical |
| `table_id` | Temporal resolution | 1hr, day, mon |
| `variable_id` | Variable short name | t2, prec, pr, tasmax |
| `grid_label` | Spatial grid | d01 (45 km), d02 (9 km), d03 (3 km) |
| `member_id` | Ensemble member | r1i1p1f1 |
| `path` | S3 Zarr store | s3://cadcat/wrf/ucla/cesm2/... |

### Models, scenarios, variables

Three helper functions summarize common catalog fields. Each takes an
`activity` of `"WRF"` or `"LOCA2"`; `NULL`, the default, returns both
activities.

#### Models

For example, list the models available for WRF:

```{r}
cae_models(activity = "WRF")
```

Or LOCA2:

```{r}
cae_models(activity = "LOCA2")
```

#### Scenarios

List available scenarios for the WRF activity:

```{r}
cae_scenarios(activity = "WRF")
```

Scenarios can also be filtered by model. In the current WRF catalog, CESM2
provides `historical`, `ssp245`, `ssp370`, and `ssp585`; ERA5 provides
`reanalysis`; most other WRF GCMs provide `historical` and `ssp370`.

```{r}
cae_scenarios(activity = "WRF", model = "CESM2")
cae_scenarios(activity = "WRF", model = "ERA5")
cae_scenarios(activity = "WRF", model = "MIROC6")
```


#### Variables

`cae_variables()` returns variable names in the catalog that match a specific
activity, timescale, or both.

For example, to see what variables are available for WRF on an hourly timescale:

```{r}
cae_variables(activity = "WRF", timescale = "1hr")
```

Available time scales include hourly, daily, and monthly (`"1hr"`, `"day"`,
`"mon"`).

#### Global search

The `cae_search()` function filters the catalog on any combination of
`activity`, `model`, `scenario`, `variable`, `timescale`, and `resolution`.

`cae_search()` combines filters with AND logic.

```{r}
cae_search(
    activity = "WRF", model = "CESM2",
    variable = "t2", timescale = "1hr", resolution = "d01"
)
```

### WRF driver-variable coverage

For WRF-to-NetCDF driver workflows, `cae_check_variables()` reports whether the
required surface meteorology variables are available or derivable for a model,
scenario, timescale, and resolution. See
[WRF Meteorological Drivers](met_drivers.qmd) for the full variable mapping and
driver-generation workflow.

```{r}
cae_check_variables("CESM2", "ssp370")
```

## Fetching

Given a combination of `activity`, `model`, `scenario`, `variable`, and time
window, `caladaptaer` reads the requested subset and returns either a `stars`
grid or a `data.frame`.

### Fetching a raster grid

The `cae_fetch()` function reads a Zarr store from S3 and returns a `stars` object.
Before reading a large dataset, test a short time window by setting
`start_time` and `end_time` in ISO 8601 format (`YYYY-MM-DD` or
`YYYY-MM-DDTHH:MM:SS`) or as `POSIXct` values. `resolution` selects the grid,
and `timescale` selects the temporal resolution, such as `"1hr"` for WRF or
`"day"` for LOCA2-Hybrid.

```{r}
t2 <- cae_fetch("t2",
    model = "CESM2", scenario = "ssp370",
    start_time = "2050-07-01T00:00:00", end_time = "2050-07-01T00:00:00"
)
t2
```

The result has three dimensions: x and y in the WRF Lambert Conformal grid and
time in UTC. Values are 2 meter air temperature in Kelvin.

#### Mapping a grid

The 45 km domain covers most of western North America. The following example
plots 2 meter air temperature on July 1, 2050 with state boundaries.

```{r}
#| fig-width: 7
#| fig-height: 7
#| fig-alt: "2 meter air temperature over the western US at 45 km on 1 July 2050 under CESM2 SSP3-7.0, with cool Sierra and Cascade ridges, warm Southwest deserts, and a cooler Pacific coast. California outlined in black."
western <- st_as_sf(
    map("state", c(
        "california", "oregon", "washington", "nevada",
        "arizona", "utah", "idaho", "montana", "wyoming",
        "colorado", "new mexico"
    ), plot = FALSE, fill = TRUE)
)
ca <- st_as_sf(map("state", "california", plot = FALSE, fill = TRUE))
western_wrf <- st_transform(western, st_crs(t2))
ca_wrf <- st_transform(ca, st_crs(t2))

t2_c <- t2
t2_c[[1]] <- units::drop_units(t2[[1]]) - 273.15

plot(t2_c,
    main = "", axes = TRUE, reset = FALSE,
    col = hcl.colors(100, "YlOrRd", rev = TRUE),
    key.pos = 4, key.width = lcm(1.5), key.length = 0.8
)
plot(st_geometry(western_wrf), add = TRUE, border = "grey30", lwd = 0.6)
plot(st_geometry(ca_wrf), add = TRUE, border = "black", lwd = 1.5)
```

To fetch data at 9 or 3 km resolution, use `resolution = "d02"` or
`resolution = "d03"`, respectively. Finer grids require more memory and longer
read times. A single 45 km timestep is approximately 0.3 MB; a single 3 km
timestep is approximately 150 MB.

### Fetching a time series from a point

`cae_fetch_point()` reads the grid and pulls the time series at the nearest
cell to a `lon`/`lat`. The transform from WGS84 to the WRF Lambert grid is
handled internally. The following example reads a summer day at Fresno:

```{r}
ts <- cae_fetch_point("t2",
    model = "CESM2", scenario = "ssp370",
    lon = -119.77, lat = 36.75,
    start_time = "2050-07-01T00:00:00",
    end_time = "2050-07-01T23:00:00"
)
head(ts)
```

This returns a `data.frame` with `time` and `value` columns that can be
plotted or analyzed directly. The `value` column for `t2` is in Kelvin. Times
are returned in UTC; the plot below converts the time axis to Pacific local
time.

```{r}
#| fig-width: 8
#| fig-height: 4.5
#| fig-alt: "Hourly 2 meter air temperature at Fresno for 1 July 2050 under CESM2 SSP3-7.0, swinging from roughly 12 to 31 degrees C across the day."
ts$temp_c <- ts$value - 273.15
ts$time_local <- as.POSIXct(
    as.numeric(ts$time),
    origin = "1970-01-01",
    tz = "America/Los_Angeles"
)

ggplot2::ggplot(ts, ggplot2::aes(x = time_local, y = temp_c)) +
    ggplot2::geom_line(linewidth = 1, color = "#D62728") +
    ggplot2::labs(x = "time (America/Los_Angeles)", y = "temperature (C)") +
    ggplot2::theme_minimal()
```

### Fetching many points at once

The `cae_fetch_points()` function reads the grid from S3 once and extracts
every site locally in a single pass. It takes a `data.frame` with `lon` and
`lat` columns and an optional `site_id`, and returns a `data.frame` with
`site_id`, `lon`, `lat`, `time`, and `value` columns.

```{r}
#| message: true
sites <- data.frame(
    site_id = c("davis", "fresno", "sacramento"),
    lon = c(-121.74, -119.77, -121.49),
    lat = c(38.54, 36.75, 38.58)
)
mp <- cae_fetch_points("t2",
    model = "CESM2", scenario = "ssp370",
    points = sites,
    start_time = "2050-07-01T00:00:00",
    end_time = "2050-07-01T05:00:00"
)
head(mp, 8)
```

### Comparing models and scenarios

Because WRF GCMs are downscaled to a common spatial and temporal resolution,
the same point-extraction workflow can be applied across models and scenarios.

The following example fetches the same day at Fresno from every WRF GCM under
SSP3-7.0 and plots the resulting time series together.

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Hourly 2 meter air temperature at Fresno for 1 July 2050 from every WRF GCM under SSP3-7.0. The curves share the cool night, warm afternoon shape but separate by a few degrees C in absolute level."
wrf_370 <- Filter(
    function(m) "ssp370" %in% cae_scenarios(activity = "WRF", model = m),
    setdiff(cae_models(activity = "WRF"), "ERA5")
)

model_ts <- lapply(wrf_370, function(m) {
    d <- cae_fetch_point("t2",
        model = m, scenario = "ssp370",
        lon = -119.77, lat = 36.75,
        start_time = "2050-07-01T00:00:00",
        end_time = "2050-07-01T23:00:00"
    )
    d$model <- m
    d$temp_c <- d$value - 273.15
    d
})
model_ts <- do.call(rbind, model_ts)

cols <- hcl.colors(length(wrf_370), "Dark 3")
plot(model_ts$time, model_ts$temp_c, type = "n", xlab = "time", ylab = "temperature (C)")
for (i in seq_along(wrf_370)) {
    d <- model_ts[model_ts$model == wrf_370[i], ]
    lines(d$time, d$temp_c, col = cols[i], lwd = 2)
}
legend("topright", legend = wrf_370, col = cols, lwd = 2, cex = 0.7, bty = "n")
```

## Where to go next

- [WRF Meteorological Drivers](met_drivers.qmd) -- WRF to CF-standard NetCDF
  meteorological drivers
- [Package reference](../reference/index.html) -- every function, every
  argument, every default

## References

- Rahimi, S. et al. (2024). An overview of the Western United States
  Dynamically Downscaled Dataset (WUS-D3). *Geoscientific Model Development*,
  17, 2265--2286.
  [doi:10.5194/gmd-17-2265-2024](https://doi.org/10.5194/gmd-17-2265-2024)
- Rahimi, S., Krantz, W., Lin, Y., Bass, B., Goldenson, N., Hall, A., Lebo, Z.,
  and Norris, J. (2022). Evaluation of a reanalysis-driven configuration of
  WRF4 over the Western United States from 1980 to 2020. *Journal of
  Geophysical Research: Atmospheres*, 127, e2021JD035699.
  [doi:10.1029/2021JD035699](https://doi.org/10.1029/2021JD035699)
- Pierce, D. W. et al. Statistical downscaling using Localized Constructed
  Analogs (LOCA). *Journal of Hydrometeorology*, 15(6), 2558--2585.
- Cal-Adapt Analytics Engine: <https://analytics.cal-adapt.org/>
- Cal-Adapt Analytics Engine, About the Data:
  <https://analytics.cal-adapt.org/data/about/>
- CADCAT on AWS: <https://registry.opendata.aws/caladapt-coproduced-climate-data/>
