---
title: "Accessing Cal-Adapt Data in R"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

This tutorial walks through the main data access functions in `caladaptaer`.
For the shortest path from installation to a time series, start with
[Getting Started](getting_started.qmd).

These functions fall into three groups:

- **discovery** (what data are available),
- **fetching** (reading grids and point time series into R), and
- **drivers** (turning WRF output into CF-standard NetCDF for ecosystem models).

All functions are documented, with function signatures, defaults, and return types
in the [package reference](../reference/index.html) - or by using the `?`
function in R (e.g. `?cae_catalog`).

## What is on the Analytics Engine

The [Cal-Adapt Analytics Engine](https://analytics.cal-adapt.org/) hosts about
1 PB of downscaled CMIP6 climate projections for California and the western US,
produced for the state's Fifth Climate Change Assessment. The data are stored
as cloud-optimized [Zarr](https://zarr.dev/) arrays on AWS S3.

Zarr stores are chunked: a reader can request only the time and spatial chunks
needed for an analysis instead of downloading a full dataset. `caladaptaer`
uses GDAL's `/vsis3/` virtual filesystem to read those public S3 stores and
returns the requested subset as a native R object. Gridded reads return
`stars` objects; point extraction helpers return `data.frame`s.

This package provides native R access to the CAE stores.
It does not require Python, a CAE API key, or CAE authentication.

Access to two downscaling products is currently implemented:

**WRF CMIP6 (WUS-D3)** is dynamically downscaled using the
Weather Research and Forecasting (WRF) model (Rahimi et al. 2024).
The WRF downscaled runs are provided at hourly time steps on three nested grids:
45 km (d01, western US), 9 km (d02, California and surroundings), and 3 km (d03,
California). The WRF CMIP6 time domain is 1980--2100: historical GCM runs cover
1980--2014, and future scenario runs cover 2015--2100. Eight General
Circulation Models (GCMs) are downscaled under the Shared Socioeconomic Pathway
SSP3-7.0 scenario. CESM2 is also available under SSP2-4.5 and SSP5-8.5.

The Analytics Engine also includes an ERA5-driven WRF historical reconstruction.
This is a reanalysis-based product, not a future GCM scenario. The underlying
WRF-ERA5 evaluation covers 1980--2020 (Rahimi et al. 2022); in the Analytics
Engine, reanalysis products are described separately from GCM historical runs.

The downscaled WRF output includes a range of surface meteorology variables. These
vary by model, but all include temperature, pressure, precipitation, humidity,
solar radiation, and wind.

**LOCA2-Hybrid**, statistically downscaled by Scripps (Pierce et al.). Daily
output at 3 km for 15 GCMs across three SSPs, covering 1950 to 2100.

### Choosing between products

The products differ in ways that matter for downstream analysis:

| Product | Time step | Time domain | Ensemble | Use when |
|---|---|---|---|---|
| WRF CMIP6 | Hourly | 1980--2014 historical GCM runs; 2015--2100 SSP runs | Smaller GCM ensemble; SSP3-7.0 is broadest, with additional CESM2 scenarios | You need sub-daily meteorology, pressure, radiation, wind, or historical/future simulations from the same downscaling framework. |
| WRF-ERA5 historical reconstruction | Hourly | 1980--2020 | Single reanalysis-driven run | You need an observation-constrained historical reconstruction rather than a future climate scenario. |
| LOCA2-Hybrid | Daily | 1950--2014 historical GCM runs; 2015--2100 SSP runs | Broader GCM and SSP coverage | You need a larger daily ensemble or daily temperature and precipitation extremes, and do not need hourly drivers. |

WRF CMIP6 provides historical and future simulations at the same spatial
resolution and with the same dynamical downscaling method, but the historical
GCM runs are still climate model simulations, not observations. The ERA5-driven
WRF reconstruction is historical only and is better suited when a reanalysis
baseline is needed.

## Setup

See [Getting Started](getting_started.qmd) for installation instructions. This
tutorial assumes `caladaptaer` and the supporting spatial packages are already
installed.

```{r}
library(caladaptaer)
library(stars)
library(sf)
library(maps)
```

The Zarr driver requires GDAL 3.4 or newer:

```{r}
sf::sf_extSoftVersion()["GDAL"]
```

## Discovery

### CAE Catalog Inventory

The `cae_catalog()` lists the CADCAT index, one row per variable per
model/scenario/resolution combo, and caches it for the session.

```{r}
cat <- cae_catalog()
nrow(cat)
names(cat)
head(cat)
```

The columns of the catalog follow CMIP6 naming. All functions use this vocabulary:

| Column | Meaning | Examples |
|--------|---------|----------|
| `activity_id` | Downscaling method | WRF, LOCA2 |
| `source_id` | Parent GCM | CESM2, MIROC6, EC-Earth3 |
| `experiment_id` | Scenario | ssp370, ssp245, historical |
| `table_id` | Temporal resolution | 1hr, day, mon |
| `variable_id` | Variable short name | t2, prec, pr, tasmax |
| `grid_label` | Spatial grid | d01 (45 km), d02 (9 km), d03 (3 km) |
| `member_id` | Ensemble member | r1i1p1f1 |
| `path` | S3 Zarr store | s3://cadcat/wrf/ucla/cesm2/... |

### Models, scenarios, variables

Three helper functions make it easy to explore the catalog.
Each takes an `activity` of `"WRF"` or `"LOCA2"` (`NULL`, the default, returns both)
and returns a character vector of the unique values in that column for that activity.


#### Models

For example, list the models available for WRF:

```{r}
cae_models(activity = "WRF")
```

Or LOCA2:

```{r}
cae_models(activity = "LOCA2")
```

#### Scenarios

List available scenarios for the WRF activity:

```{r}
cae_scenarios(activity = "WRF")
```

You can also filter scenarios by model. Note that WRF-downscaled CESM2 only provides 'historical' and three SSPs; ERA5 provides only 'historical'; and the other WRF GCMs provide only 'historical and one SSP':

```{r}
cae_scenarios(activity = "WRF", model = "CESM2")
cae_scenarios(activity = "WRF", model = "ERA5")
cae_scenarios(activity = "WRF", model = "MIROC6")
```


#### Variables

The `cae_variables()` function returns a character vector of variable names in the catalog that match a specific activity and / or timescale.

For example, to see what variables are available for WRF on an hourly timescale:

```{r}
cae_variables(activity = "WRF", timescale = "day")
```

Available time scales include hourly, daily, and monthly (`"1hr"`, `"day"`, `"mon"`).

#### Global Search

The `cae_search()` function filters the catalog on any combination of `activity`, `model`,
`scenario`, `variable`, `timescale`, and `resolution`.

The filter uses AND to combine the arguments and filter the catalog.

```{r}
cae_search(
    activity = "WRF", model = "CESM2",
    variable = "t2", timescale = "1hr", resolution = "d01"
)
```

### Checking a model driver set before you fetch

We have provided a helper function to confirm the
availability of a core set of variables that are commonly used as drivers for ecosystem models.

| WRF name | Field | Units |
|----------|-------|-------|
| `t2` | 2 meter air temperature | K |
| `psfc` | Surface pressure | Pa |
| `prec` | Precipitation | mm/timestep |
| `q2` | Water vapor mixing ratio | kg/kg |
| `swdnb` | Downward shortwave radiation | W/m2 |
| `u10` | 10 meter eastward wind | m/s |
| `v10` | 10 meter northward wind | m/s |

Before you commit to building met drivers `cae_check_variables()`
will report whether each of the eight fields wants is directly available or can be derived from other variables.

```{r}
cae_check_variables("CESM2", "ssp370")
```

## Data Access

Given a combination of `activity`, `model`, `scenario`, `variable`, and time window,
`caladaptaer` reads the requested subset and returns either a `stars` grid or a
`data.frame`. The next sections show how to fetch both grids and point time series.

### Fetching a Raster Grid

The `cae_fetch()` function reads a Zarr store from S3 and returns a `stars` object.
Before reading in a large dataset, test a short time window by setting the `start_time` and `end_time` either in the ISO 8601
time format (YYYY-MM-DD or YYYY-MM-DDTHH:MM:SS)
or POSIXct format.
The  `resolution` picks the grid and `timescale` picks `1hr` (WRF) or `day` (LOCA2).

```{r}
t2 <- cae_fetch("t2",
    model = "CESM2", scenario = "ssp370",
    start_time = "2050-07-01T00:00:00", end_time = "2050-07-01T00:00:00"
)
t2
```

The result has three dimensions: x and y in the WRF Lambert Conformal grid and
time in UTC. Values are 2 meter air temperature in Kelvin.

#### Mapping a Grid

The 45 km domain covers most of western North America. Next, we will
demonstrate how to plot the map of 2m air temperature on July 1, 2050 with state boundaries.

```{r}
#| fig-width: 7
#| fig-height: 7
#| fig-alt: "2 meter air temperature over the western US at 45 km on 1 July 2050 under CESM2 SSP3-7.0, with cool Sierra and Cascade ridges, warm Southwest deserts, and a cooler Pacific coast. California outlined in black."
western <- st_as_sf(
    map("state", c(
        "california", "oregon", "washington", "nevada",
        "arizona", "utah", "idaho", "montana", "wyoming",
        "colorado", "new mexico"
    ), plot = FALSE, fill = TRUE)
)
ca <- st_as_sf(map("state", "california", plot = FALSE, fill = TRUE))
western_wrf <- st_transform(western, st_crs(t2))
ca_wrf <- st_transform(ca, st_crs(t2))

t2_c <- t2
t2_c[[1]] <- units::drop_units(t2[[1]]) - 273.15

plot(t2_c,
    main = "", axes = TRUE, reset = FALSE,
    col = hcl.colors(100, "YlOrRd", rev = TRUE),
    key.pos = 4, key.width = lcm(1.5), key.length = 0.8
)
plot(st_geometry(western_wrf), add = TRUE, border = "grey30", lwd = 0.6)
plot(st_geometry(ca_wrf), add = TRUE, border = "black", lwd = 1.5)
```

To fetch data with 9 or 3 km resolution, use `resolution = "d02"` or `"d03"`, respectively.
But be aware that the finer grids are much larger, so they will take longer to read and use more memory - a timestep at 45km is approximately 0.3MB while and is 150 MB at 3 km.

### Fetching a Time Series from a Point

`cae_fetch_point()` reads the grid and pulls the time series at the nearest
cell to a `lon`/`lat`. The transform from WGS84 to the WRF Lambert grid is
handled for you. Here is a summer day at Fresno:

```{r}
ts <- cae_fetch_point("t2",
    model = "CESM2", scenario = "ssp370",
    lon = -119.77, lat = 36.75,
    start_time = "2050-07-01T00:00:00",
    end_time = "2050-07-01T23:00:00"
)
head(ts)
```

This returns a `data.frame` with `time` and `value` columns that can be
plotted or analyzed directly. Recall that the `value` column for `t2` is in Kelvin, and time is in UTC; we will convert these before plotting.

```{r}
#| fig-width: 8
#| fig-height: 4.5
#| fig-alt: "Hourly 2 meter air temperature at Fresno for 1 July 2050 under CESM2 SSP3-7.0, swinging from roughly 12 to 31 degrees C across the day."
ts$temp_c <- ts$value - 273.15
ts$time <- lubridate::with_tz(ts$time, tz = "PST")
plot(ts$time, ts$temp_c,
    type = "l", lwd = 2, col = "#D62728",
    xlab = "time", ylab = "temperature (C)"
)
```

### Fetching Many points at Once

The `cae_fetch_points()` function reads the grid from S3 once, and then extracts
every site locally in a single pass. It takes a `data.frame` with `lon` and `lat`
columns (and an optional `site_id`) and returns a `data.frame` with `site_id`, `lon`, `lat`, `time`, and `value` columns.

```{r}
#| message: true
sites <- data.frame(
    site_id = c("davis", "fresno", "sacramento"),
    lon = c(-121.74, -119.77, -121.49),
    lat = c(38.54, 36.75, 38.58)
)
mp <- cae_fetch_points("t2",
    model = "CESM2", scenario = "ssp370",
    points = sites,
    start_time = "2050-07-01T00:00:00",
    end_time = "2050-07-01T05:00:00"
)
head(mp, 8)
```

### Comparing Models and Scenarios

Because every GCM is downscaled to a common spatial and temporal resolution,  it is straightforward to evaluate differences across models and scenarios.

Here we fetch the same day at Fresno from every WRF GCM under SSP3-7.0, and plot them together. The curves share the cool night, warm afternoon shape but separate by a few degrees C in absolute level.

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Hourly 2 meter air temperature at Fresno for 1 July 2050 from every WRF GCM under SSP3-7.0. The curves share the cool night, warm afternoon shape but separate by a few degrees C in absolute level."
wrf_370 <- Filter(
    function(m) "ssp370" %in% cae_scenarios(activity = "WRF", model = m),
    setdiff(cae_models(activity = "WRF"), "ERA5")
)

model_ts <- lapply(wrf_370, function(m) {
    d <- cae_fetch_point("t2",
        model = m, scenario = "ssp370",
        lon = -119.77, lat = 36.75,
        start_time = "2050-07-01T00:00:00",
        end_time = "2050-07-01T23:00:00"
    )
    d$model <- m
    d$temp_c <- d$value - 273.15
    d
})
model_ts <- do.call(rbind, model_ts)

cols <- hcl.colors(length(wrf_370), "Dark 3")
plot(model_ts$time, model_ts$temp_c, type = "n", xlab = "time", ylab = "temperature (C)")
for (i in seq_along(wrf_370)) {
    d <- model_ts[model_ts$model == wrf_370[i], ]
    lines(d$time, d$temp_c, col = cols[i], lwd = 2)
}
legend("topright", legend = wrf_370, col = cols, lwd = 2, cex = 0.7, bty = "n")
```

## Where to go next

- [SIPNET Met Drivers](sipnet_met_drivers.qmd) -- raw WRF to CF standard NetCDF
- [Bulk Met Drivers](bulk_met_drivers.qmd) -- the multi site driver path
- [Package reference](../reference/index.html) -- every function, every
  argument, every default

## References

- Rahimi, S. et al. (2024). An overview of the Western United States
  Dynamically Downscaled Dataset (WUS-D3). *Geoscientific Model Development*,
  17, 2265--2286.
  [doi:10.5194/gmd-17-2265-2024](https://doi.org/10.5194/gmd-17-2265-2024)
- Rahimi, S., Krantz, W., Lin, Y., Bass, B., Goldenson, N., Hall, A., Lebo, Z.,
  and Norris, J. (2022). Evaluation of a reanalysis-driven configuration of
  WRF4 over the Western United States from 1980 to 2020. *Journal of
  Geophysical Research: Atmospheres*, 127, e2021JD035699.
  [doi:10.1029/2021JD035699](https://doi.org/10.1029/2021JD035699)
- Pierce, D. W. et al. Statistical downscaling using Localized Constructed
  Analogs (LOCA). *Journal of Hydrometeorology*, 15(6), 2558--2585.
- Cal-Adapt Analytics Engine: <https://analytics.cal-adapt.org/>
- Cal-Adapt Analytics Engine, About the Data:
  <https://analytics.cal-adapt.org/data/about/>
- CADCAT on AWS: <https://registry.opendata.aws/caladapt-coproduced-climate-data/>
