---
title: "Getting Started with caladaptR"
---

```{r}
#| include: false
knitr::opts_chunk$set(eval = FALSE)
```

::: {.callout-note}
## What you will learn

- What climate data is available on the Cal-Adapt Analytics Engine and how the
  CADCAT catalog organizes it
- How to query the catalog to find datasets by model, scenario, variable, and
  resolution
- How to read gridded Zarr data from AWS S3 directly into R as a `stars` object
- How to produce California climate maps comparable to the official Python
  notebooks
- How to extract and visualize time series at specific locations
:::

## The Cal-Adapt Analytics Engine

The [Cal-Adapt Analytics Engine](https://analytics.cal-adapt.org/) hosts
roughly 1 PB of downscaled CMIP6 climate projections for California and the
western US, produced for the state's Fifth Climate Change Assessment. The
data lives in cloud-optimized [Zarr](https://zarr.dev/) stores on AWS S3,
publicly accessible with no credentials required.

Two downscaling products are available:

**WRF CMIP6 (WUS-D3)** -- dynamically downscaled by UCLA's Center for Climate
Science (Rahimi et al. 2024). The Weather Research and Forecasting model
produces hourly output at three nested resolutions: 45 km (d01, western US),
9 km (d02, California + surroundings), and 3 km (d03, California). Eight
GCMs are downscaled, primarily under SSP3-7.0, with CESM2 also available
for SSP2-4.5 and SSP5-8.5. The WRF output includes the full surface
meteorology -- temperature, pressure, precipitation, humidity, radiation,
and wind components.

**LOCA2-Hybrid** -- statistically downscaled by Scripps/UCSD (Pierce et al.).
Daily output at 3 km for 15 GCMs across three SSPs (SSP2-4.5, SSP3-7.0,
SSP5-8.5), covering 1950--2100. LOCA2-Hybrid uses a hybrid analog approach
that substitutes WRF-derived weather patterns for end-of-century conditions
where the historical analog library runs thin.

The official Python interface is
[climakitae](https://github.com/cal-adapt/climakitae). `caladaptR` provides
the same data access natively in R, reading Zarr stores via GDAL's `/vsis3/`
virtual filesystem. No Python, no API key, no authentication.

## Setup

```{r}
library(caladaptR)
library(stars)
library(sf)
library(ggplot2)
library(maps)
```

Check that GDAL supports Zarr (version 3.4+):

```{r}
sf::sf_extSoftVersion()["GDAL"]
```

## Exploring the catalog {#sec-catalog}

The CADCAT catalog is a CSV index of every Zarr store on S3. Each row is one
variable from one model/scenario/resolution combination. `ca_catalog()`
downloads the table once per session and caches it.

```{r}
cat <- ca_catalog()
nrow(cat)
names(cat)
```

The catalog follows CMIP6 naming conventions:

| Column | Description | Examples |
|--------|-------------|---------|
| `activity_id` | Downscaling method | WRF, LOCA2 |
| `source_id` | Parent GCM | CESM2, MIROC6, EC-Earth3 |
| `experiment_id` | Scenario | ssp370, ssp245, historical |
| `table_id` | Temporal resolution | 1hr, day, mon |
| `variable_id` | Variable short name | t2, prec, pr, tasmax |
| `grid_label` | Spatial resolution | d01 (45 km), d02 (9 km), d03 (3 km) |
| `member_id` | Ensemble member | r1i1p1f1 |
| `path` | S3 Zarr store path | s3://cadcat/wrf/ucla/cesm2/... |

### Data availability

The WRF and LOCA2 datasets differ in their model/scenario coverage. Use the
convenience functions to see what's available:

```{r}
# WRF GCMs -- 8 models dynamically downscaled
ca_models(activity = "WRF")

# LOCA2 GCMs -- 15 models statistically downscaled
ca_models(activity = "LOCA2")

# CESM2 has the broadest scenario coverage in WRF
ca_scenarios(activity = "WRF", model = "CESM2")

# hourly WRF variables
ca_variables(activity = "WRF", timescale = "1hr")
```

Here's the full picture of what's on S3 for the WRF hourly data at 45 km:

![WRF data availability across GCMs and scenarios. CESM2 has the broadest
coverage with three SSPs; all other models have SSP3-7.0 and
historical.](figures/fig07_catalog_overview.png){#fig-catalog width=90%}

::: {.callout-tip}
## WRF vs LOCA2 -- when to use which

Use **WRF** when you need hourly data, process-based meteorology (radiation,
humidity, pressure), or when you care about mesoscale features like orographic
precipitation or sea breeze effects. Use **LOCA2** when you need more GCMs,
more SSPs, or daily temperature/precipitation extremes at the finest
available resolution (3 km).
:::

### Searching the catalog

`ca_search()` filters by any combination of fields. All filters are ANDed.

```{r}
# all hourly CESM2 temperature at 45 km
ca_search(activity = "WRF", model = "CESM2",
          variable = "t2", timescale = "1hr", resolution = "d01")
```


## Fetching gridded data {#sec-fetch}

`ca_fetch()` reads a Zarr store from S3 and returns a `stars` object. Start
with `n_timesteps = 1` to keep the initial read small -- a single 45 km
timestep is about 0.3 MB.

```{r}
t2 <- ca_fetch("t2", model = "CESM2", scenario = "ssp370",
               n_timesteps = 1)
t2
```

The result has three dimensions: x and y in the WRF Lambert Conformal Conic
projection, and time in UTC. The values are 2-meter air temperature in
Kelvin.

### Mapping temperature over the western US

The 45 km WRF domain covers most of western North America. Overlay state
boundaries to orient the map -- this is the R equivalent of what the
Cal-Adapt Python notebooks do with cartopy.

```{r}
# state boundaries for overlay
ca_sf <- st_as_sf(map("state", "california", plot = FALSE, fill = TRUE))
western_sf <- st_as_sf(
  map("state", c("california", "oregon", "washington", "nevada",
                  "arizona", "utah", "idaho", "montana", "wyoming",
                  "colorado", "new mexico"), plot = FALSE, fill = TRUE)
)

# transform boundaries to the WRF projection
ca_wrf <- st_transform(ca_sf, st_crs(t2))
ws_wrf <- st_transform(western_sf, st_crs(t2))

# convert to Celsius and plot
t2_c <- t2
t2_c[[1]] <- units::drop_units(t2[[1]]) - 273.15

plot(t2_c, main = "", axes = TRUE, reset = FALSE,
     col = hcl.colors(100, "YlOrRd", rev = TRUE),
     key.pos = 4, key.width = lcm(1.5), key.length = 0.8)
plot(st_geometry(ws_wrf), add = TRUE, border = "grey30", lwd = 0.6)
plot(st_geometry(ca_wrf), add = TRUE, border = "black", lwd = 1.5)
title("2-meter Air Temperature (C)\nCESM2 SSP3-7.0, WRF 45 km")
```

![WRF 2-meter air temperature over the western US at 45 km resolution. The
spatial pattern shows expected features -- cool mountain temperatures in the
Sierra Nevada and Cascades, warm desert conditions in the Southwest, and a
maritime gradient along the Pacific coast. California boundary shown in
black.](figures/fig01_wrf_t2_western_us.png){#fig-t2-map width=80%}

### Shortwave radiation

Surface downwelling shortwave radiation drives photosynthesis in ecosystem
models and solar generation in energy applications. The WRF field captures
the diurnal solar cycle and cloud effects:

```{r}
sw <- ca_fetch("swdnb", model = "CESM2", scenario = "ssp370",
               n_timesteps = 1)
```

![Shortwave radiation at the surface. The day-night terminator is visible
as a sharp gradient. Desert regions receive the highest clear-sky irradiance.
Cloud shadows appear as dark patches in otherwise sunny
areas.](figures/fig03_wrf_swdown.png){#fig-sw width=80%}


### Surface pressure

Surface pressure is a required input for ecosystem models and reveals the
topographic structure of the domain:

```{r}
psfc <- ca_fetch("psfc", model = "CESM2", scenario = "ssp370",
                 n_timesteps = 1)
```

![Surface pressure (hPa) over the western US. The topographic imprint is
immediately visible -- low pressure over the Rocky Mountains, Sierra Nevada,
and Cascades where elevation reduces the overlying atmospheric column, and
higher pressure over the Central Valley and coastal lowlands. This is one of
the 7 meteorological variables SIPNET needs for carbon flux
simulations.](figures/fig04_wrf_psfc.png){#fig-psfc width=80%}


## Point extraction {#sec-point}

For single-site analysis, `ca_fetch_point()` reads the grid and extracts
the time series at the nearest grid cell to a lon/lat coordinate. CRS
transformation from WGS84 to the WRF Lambert grid is handled automatically.

```{r}
# 24 hours of temperature at Fresno, CA
ts <- ca_fetch_point("t2", model = "CESM2", scenario = "ssp370",
                     lon = -119.77, lat = 36.75,
                     n_timesteps = 24)
head(ts)
```

```{r}
ts$temp_c <- ts$value - 273.15
plot(ts$time, ts$temp_c, type = "l", lwd = 2, col = "#D62728",
     xlab = "Time (UTC)", ylab = "Temperature (C)",
     main = "Diurnal Temperature -- Fresno, CA")
points(ts$time, ts$temp_c, pch = 16, cex = 0.6, col = "#D62728")
grid(col = "grey85")
```

![Diurnal temperature cycle at Fresno extracted from the CESM2 SSP3-7.0
WRF simulation. The Central Valley in September shows cool overnight lows
near 12 C and afternoon highs above 30 C -- a reasonable diurnal range for
the location and season.](figures/fig02_fresno_diurnal.png){#fig-diurnal width=80%}


## Multi-model comparison {#sec-multimodel}

One of the strengths of the Cal-Adapt dataset is that multiple GCMs are
available under the same downscaling framework. Comparing models at the
same location reveals inter-model spread -- a measure of structural
uncertainty in the projections.

```{r}
# get all WRF models with SSP3-7.0 (excluding ERA5 reanalysis)
wrf_models <- setdiff(ca_models(activity = "WRF"), "ERA5")
models_370 <- character(0)
for (m in wrf_models) {
  if ("ssp370" %in% ca_scenarios(activity = "WRF", model = m))
    models_370 <- c(models_370, m)
}

# extract 24-hour temperature at Fresno from each model
model_ts <- lapply(models_370, function(m) {
  d <- ca_fetch_point("t2", model = m, scenario = "ssp370",
                      lon = -119.77, lat = 36.75, n_timesteps = 24)
  d$model <- m
  d$temp_c <- d$value - 273.15
  d
})
model_ts <- do.call(rbind, model_ts)
```

![Diurnal temperature from all WRF GCMs at Fresno under SSP3-7.0. The
models agree on the general shape -- cool night, warm afternoon -- but
differ by 2--5 C in the absolute magnitude. This spread reflects
structural differences in how each GCM represents the large-scale
circulation that drives the WRF boundary
conditions.](figures/fig06_multimodel_fresno.png){#fig-multimodel width=90%}


## Multiple meteorological variables {#sec-met}

Ecosystem models and energy system models typically need several
meteorological inputs. The WRF hourly dataset includes the full surface
meteorology. Here we extract all 7 SIPNET-required variables at a point:

| WRF name | Description | Units |
|----------|-------------|-------|
| `t2` | 2-meter air temperature | K |
| `psfc` | Surface pressure | Pa |
| `prec` | Accumulated precipitation | mm/timestep |
| `q2` | Water vapor mixing ratio | kg/kg |
| `swdnb` | Downward shortwave radiation | W/m2 |
| `u10` | 10-meter eastward wind | m/s |
| `v10` | 10-meter northward wind | m/s |

```{r}
vars <- c("t2", "swdnb", "prec", "q2", "psfc", "u10", "v10")
pts <- lapply(vars, function(v) {
  ca_fetch_point(v, model = "CESM2", scenario = "ssp370",
                 lon = -119.77, lat = 36.75, n_timesteps = 24)
})
names(pts) <- vars
```

![Six-panel display of hourly meteorological variables at Fresno, CA from
the CESM2 SSP3-7.0 WRF simulation. Top row: air temperature (C) and
surface shortwave radiation (W/m2). Middle row: precipitation flux
(kg/m2/s, converted from mm/timestep) and specific humidity (g/kg,
converted from mixing ratio). Bottom row: wind speed (m/s, computed from
U and V components) and surface pressure (hPa). The diurnal signatures
are physically consistent -- peak radiation drives peak temperature with
a lag, humidity drops as temperature rises, and pressure follows a
semidiurnal tide.](figures/fig05_met_panel_fresno.png){#fig-met width=95%}


## What's next

- **Writing CF-standard NetCDF:** See the
  [SIPNET Met Drivers](sipnet_met_drivers.qmd) vignette for the full pipeline
  from raw WRF variables to CF-compliant met files.

- **LOCA2 daily data:** The same `ca_fetch()` interface works for LOCA2 --
  just specify `timescale = "day"` and a LOCA2 model name (e.g. `"EC-Earth3"`).

- **Spatial subsetting:** For California-only analysis at 9 or 3 km, use
  `resolution = "d02"` or `"d03"`, but watch memory -- a single 3 km timestep
  is ~150 MB.


## Requirements

- **R >= 4.1** (native pipe support)
- **GDAL >= 3.4** (Zarr driver). Check with `sf::sf_extSoftVersion()`.
- **Internet access** to AWS S3 (us-west-2 region). No authentication.
- **Memory:** The 45 km grid is 199 x 199. One timestep is ~0.3 MB. A full
  year for one variable is ~2.6 GB.


## References

- Rahimi, S. et al. (2024). An overview of the Western United States
  Dynamically Downscaled Dataset (WUS-D3). *Geoscientific Model Development*,
  17, 2265--2286.
  [doi:10.5194/gmd-17-2265-2024](https://doi.org/10.5194/gmd-17-2265-2024)
- Pierce, D. W. et al. Statistical downscaling using Localized Constructed
  Analogs (LOCA). *Journal of Hydrometeorology*, 15(6), 2558--2585.
- Krantz, W. et al. (2022). GCM skill evaluation for the Fifth California
  Climate Assessment.
- Cal-Adapt Analytics Engine: <https://analytics.cal-adapt.org/>
- CADCAT on AWS: <https://registry.opendata.aws/caladapt-coproduced-climate-data/>
