---
title: "Getting Started"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

caladaptaer reads downscaled CMIP6 climate projections off the
[Cal-Adapt Analytics Engine](https://analytics.cal-adapt.org/) directly into
R. Data lives as cloud optimized Zarr on AWS S3, and the package reads it
through GDAL: no Python, no API key, no authentication. Every call hands back
a native R object, a `stars` grid or a tidy `data.frame`, so results drop
right into the R geospatial and modeling stack.

This page goes from install to a real time series in a few calls, then shows
the design choice that lets the same code scale to thousands of sites. Every
output below comes from actually running the code on this page. For the full
function tour see [Accessing Cal-Adapt Data in R](data_access.qmd).

## Install

```{r}
#| eval: false
remotes::install_github("lebauerapproach/caladaptaer")
```

You need R >= 4.1 and GDAL >= 3.4 for the Zarr driver. Confirm your GDAL is
recent enough:

```{r}
sf::sf_extSoftVersion()["GDAL"]
```

```{r}
library(caladaptaer)
```

## Discover what's available

Ask what the WRF dynamical downscaling covers:

```{r}
cae_models(activity = "WRF")
```

Eight GCMs plus the ERA5 reanalysis driven run. CESM2 carries the widest set
of scenarios:

```{r}
cae_scenarios(activity = "WRF", model = "CESM2")
```

## Pull a time series at a point

`cae_fetch_point()` reads the grid and extracts the nearest cell to a
lon/lat. Transform from WGS84 to the WRF Lambert Conformal grid happens for
you. Here's a summer day in 2050 at Fresno under SSP3-7.0:

```{r}
ts <- cae_fetch_point("t2", model = "CESM2", scenario = "ssp370",
                      lon = -119.77, lat = 36.75,
                      start_time = "2050-07-01T00:00:00",
                      end_time   = "2050-07-01T23:00:00")
head(ts)
```

Result is a plain `data.frame` with `time` and `value` in Kelvin, so base R
plotting just works. Times are UTC, so the midday dip is the local overnight
low (UTC noon is roughly 5am Pacific).

```{r}
#| fig-width: 8
#| fig-height: 4.5
#| fig-alt: "Hourly 2 meter air temperature at Fresno for 1 July 2050 under CESM2 SSP3-7.0, ranging from about 12 to 31 degrees C across the day."
ts$temp_c <- ts$value - 273.15
plot(ts$time, ts$temp_c, type = "l", lwd = 2, col = "#D62728",
     xlab = "time", ylab = "temperature (C)")
```

## Extract many locations in one read

This is the design choice worth knowing up front. `cae_fetch_points()` reads
the grid from S3 once and extracts every location locally in a single
vectorized pass. Three sites or ten thousand, it's one S3 read per variable,
not one per site.

```{r}
#| message: true
sites <- data.frame(
  site_id = c("davis", "fresno", "sacramento"),
  lon = c(-121.74, -119.77, -121.49),
  lat = c(38.54, 36.75, 38.58)
)
mp <- cae_fetch_points("t2", model = "CESM2", scenario = "ssp370",
                       points = sites,
                       start_time = "2050-07-01T00:00:00",
                       end_time   = "2050-07-01T05:00:00")
head(mp, 8)
```

## Where to go next

- [Accessing Cal-Adapt Data in R](data_access.qmd) -- every function with
  output: catalog search, gridded fetch, California maps, multi model
  comparison, and the full surface meteorology
- [SIPNET Met Drivers](sipnet_met_drivers.qmd) -- raw WRF to CF standard
  NetCDF for ecosystem models
- [Bulk Met Drivers](bulk_met_drivers.qmd) -- the multi site extraction path
