---
title: "Getting Started"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`caladaptaer` provides access to downscaled CMIP6 climate projections from the
[Cal-Adapt Analytics Engine](https://analytics.cal-adapt.org/). This tutorial
is the shortest path from installation to reading and plotting a time series in
R.

For a more comprehensive exposition of the functions, see
[Accessing Cal-Adapt Data in R](data_access.qmd).

## Installation

Requirements:

- R >= 4.1
- GDAL >= 3.4
- internet connection
- R packages:
  - `sf`, `stars`, `dplyr`, `ggplot2` (and their dependencies)
  - `remotes` (for installation from GitHub)

```{r}
#| eval: false
remotes::install_github("lebauerapproach/caladaptaer")
```

To confirm that you have a compatible version of GDAL, check the version:

```{r}
sf::sf_extSoftVersion()["GDAL"]
```

```{r}
library(caladaptaer)
```

## Pull a time series at a point

`cae_fetch_point()` reads the grid and extracts data from the grid cell that is
closest to the provided longitude and latitude.

Here's a summer day in Fresno in 2050 under SSP3-7.0:

```{r}
ts <- cae_fetch_point("t2",
    model = "CESM2", scenario = "ssp370",
    lon = -119.77, lat = 36.75,
    start_time = "2050-07-01T00:00:00",
    end_time = "2050-07-01T23:00:00"
)
head(ts)
```

The result is a `data.frame` with `time` and `value` in Kelvin, so base R
plotting just works. Times are UTC, so the midday dip is the local overnight
low (UTC noon is roughly 5am Pacific).

```{r}
#| fig-width: 8
#| fig-height: 4.5
#| fig-alt: "Hourly 2 meter air temperature at Fresno for 1 July 2050 under CESM2 SSP3-7.0, ranging from about 12 to 31 degrees C across the day."
ts$temp_c <- ts$value - 273.15
plot(ts$time, ts$temp_c,
    type = "l", lwd = 2, col = "#D62728",
    xlab = "time", ylab = "temperature (C)"
)
```

## Extract many locations in one read

It is also possible to extract data from multiple locations in a single read
using `cae_fetch_points()`. This function reads the grid from S3 once and
extracts every location locally.

```{r}
#| message: true
sites <- data.frame(
    site_id = c("davis", "fresno", "sacramento"),
    lon = c(-121.74, -119.77, -121.49),
    lat = c(38.54, 36.75, 38.58)
)
mp <- cae_fetch_points("t2",
    model = "CESM2", scenario = "ssp370",
    points = sites,
    start_time = "2050-07-01T00:00:00",
    end_time = "2050-07-01T05:00:00"
)
head(mp, 8)
```

## Where to go next

- [Accessing Cal-Adapt Data in R](data_access.qmd) -- every function with
  output: catalog search, gridded fetch, California maps, multi model
  comparison, and the full surface meteorology
- [SIPNET Met Drivers](sipnet_met_drivers.qmd) -- raw WRF to CF standard
  NetCDF for ecosystem models
- [Bulk Met Drivers](bulk_met_drivers.qmd) -- the multi site extraction path
