---
title: "Getting Started"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`caladaptaer` provides access to downscaled CMIP6 climate projections from the
[Cal-Adapt Analytics Engine](https://analytics.cal-adapt.org/). This tutorial
covers installation, catalog checks, and the first point time-series read.

For a broader data-access guide, see
[Accessing Cal-Adapt Data in R](data_access.qmd).

## Installation

Requirements:

- R >= 4.1
- GDAL >= 3.4
- internet connection
- R packages:
  - `sf`, `stars`, `dplyr`, `ggplot2` (and their dependencies)
  - `remotes` (for installation from GitHub)

```{r}
#| eval: false
remotes::install_github("lebauerapproach/caladaptaer")
```

Check the GDAL version before reading from the cloud-hosted Zarr stores:

```{r}
sf::sf_extSoftVersion()["GDAL"]
```

```{r}
library(caladaptaer)
```

## Fetch a time series at a point

`cae_fetch_point()` reads the grid and extracts data from the grid cell that is
closest to the provided longitude and latitude.

The following example reads one summer day in Fresno in 2050 under SSP3-7.0:

```{r}
ts <- cae_fetch_point("t2",
    model = "CESM2", scenario = "ssp370",
    lon = -119.77, lat = 36.75,
    start_time = "2050-07-01T00:00:00",
    end_time = "2050-07-01T23:00:00"
)
head(ts)
```

The result is a `data.frame` with `time` and `value` columns. Values for `t2`
are in Kelvin. Times are returned in UTC; the plot below converts the time axis
to Pacific local time.

```{r}
#| fig-width: 8
#| fig-height: 4.5
#| fig-alt: "Hourly 2 meter air temperature at Fresno for 1 July 2050 under CESM2 SSP3-7.0, ranging from about 12 to 31 degrees C across the day."
ts$temp_c <- ts$value - 273.15
ts$time_local <- as.POSIXct(
    as.numeric(ts$time),
    origin = "1970-01-01",
    tz = "America/Los_Angeles"
)

ggplot2::ggplot(ts, ggplot2::aes(x = time_local, y = temp_c)) +
    ggplot2::geom_line(linewidth = 1, color = "#D62728") +
    ggplot2::labs(x = "time (America/Los_Angeles)", y = "temperature (C)") +
    ggplot2::theme_minimal()
```

## Extract many locations in one read

Use `cae_fetch_points()` to extract data from multiple locations in a single
read. This function reads the grid from S3 once and extracts every location
locally.

```{r}
#| message: true
sites <- data.frame(
    site_id = c("davis", "fresno", "sacramento"),
    lon = c(-121.74, -119.77, -121.49),
    lat = c(38.54, 36.75, 38.58)
)
mp <- cae_fetch_points("t2",
    model = "CESM2", scenario = "ssp370",
    points = sites,
    start_time = "2050-07-01T00:00:00",
    end_time = "2050-07-01T05:00:00"
)
head(mp, 8)
```

## Where to go next

- [Accessing Cal-Adapt Data in R](data_access.qmd) -- dataset background,
  catalog search, gridded fetch, point extraction, and model comparison
- [WRF Meteorological Drivers](met_drivers.qmd) -- WRF to CF-standard NetCDF
  meteorological drivers for site-based model workflows
