---
title: "WRF Meteorological Drivers"
---

```{r}
#| include: false
knitr::opts_chunk$set(eval = FALSE)
```

## Purpose

This guide describes how to convert hourly WRF output from the Cal-Adapt
Analytics Engine into CF-1.8 NetCDF meteorological driver files. The workflow
is intended for site-based ecosystem models and other tools that consume
CF-standard meteorological forcing.

`caladaptaer` provides two driver-generation paths:

- `cae_write_cf_netcdf()` writes a CF NetCDF file from a named list of WRF
  `stars` grids.
- `cae_build_met_drivers()` extracts WRF data for multiple sites and writes one
  NetCDF file per site per year.

## Prerequisites

Install `caladaptaer` and verify that GDAL is available as described in
[Getting Started](getting_started.qmd). This guide assumes the package is
already installed.

```{r}
library(caladaptaer)
```

The driver workflow requires WRF hourly data. LOCA2-Hybrid daily products do
not provide the full hourly meteorological variable set required by these
functions.

## Required Inputs

Multi-site driver generation requires a site table with three columns:

| Column | Description |
|---|---|
| `site_id` | Site identifier used in output paths |
| `lon` | Longitude in WGS84 decimal degrees |
| `lat` | Latitude in WGS84 decimal degrees |

Additional columns can be present; `cae_build_met_drivers()` uses only
`site_id`, `lon`, and `lat`.

```{r}
sites <- data.frame(
    site_id = c("fresno", "los_angeles", "sacramento"),
    lon = c(-119.77, -118.24, -121.49),
    lat = c(36.75, 34.05, 38.58)
)
```

Driver generation also requires a WRF model, scenario, year range, spatial
grid, and output directory. Use `cae_models()`, `cae_scenarios()`, and
`cae_search()` to confirm availability before starting a large job.

## Required Variables

The CF driver files are built from eight WRF surface meteorology variables.
`wind_speed` is derived from `u10` and `v10`.

| CF variable | WRF variable | WRF units | CF units | Conversion |
|---|---|---|---|---|
| `air_temperature` | `t2` | K | K | none |
| `air_pressure` | `psfc` | Pa | Pa | none |
| `precipitation_flux` | `prec` | mm per timestep | kg m-2 s-1 | divide by timestep seconds |
| `specific_humidity` | `q2` | kg kg-1 mixing ratio | 1 | `q2 / (1 + q2)` |
| `surface_downwelling_shortwave_flux_in_air` | `swdnb` | W m-2 | W m-2 | none |
| `surface_downwelling_longwave_flux_in_air` | `lwdnb` | W m-2 | W m-2 | none |
| `eastward_wind` | `u10` | m s-1 | m s-1 | none |
| `northward_wind` | `v10` | m s-1 | m s-1 | none |
| `wind_speed` | `u10`, `v10` | m s-1 | m s-1 | `sqrt(u10^2 + v10^2)` |

Use `cae_check_variables()` to check whether the required variables are present
or derivable for a model, scenario, timescale, and resolution.

```{r}
cae_check_variables(
    model = "CESM2",
    scenario = "ssp370",
    timescale = "1hr",
    resolution = "d01"
)
```

Some WRF models store precipitation as accumulated convective and
non-convective components rather than the precomputed `prec` field.
`caladaptaer` derives `prec` from those components when possible.

## Single-Grid Workflow

Use `cae_write_cf_netcdf()` when the intended output is a CF NetCDF file for a
single WRF grid subset. The input is a named list of `stars` objects, one per
required WRF variable.

```{r}
wrf_vars <- c("t2", "psfc", "prec", "q2", "swdnb", "lwdnb", "u10", "v10")

var_data <- lapply(wrf_vars, function(variable) {
    cae_fetch(
        variable = variable,
        model = "CESM2",
        scenario = "ssp370",
        timescale = "1hr",
        resolution = "d01",
        start_time = "2050-07-01T00:00:00",
        end_time = "2050-07-01T23:00:00"
    )
})
names(var_data) <- wrf_vars

cae_write_cf_netcdf(
    var_list = var_data,
    outfile = "CESM2.ssp370.2050-07-01.cf.nc",
    dt_seconds = 3600,
    overwrite = TRUE
)
```

The output file contains CF-1.8 global metadata, CF variable names, converted
units, and derived wind speed.

## Multi-Site Workflow

Use `cae_build_met_drivers()` when the intended output is one NetCDF file per
site per year. The function reads each WRF variable once per year, extracts all
sites from that grid, and writes site-year files.

```{r}
result <- cae_build_met_drivers(
    sites = sites,
    model = "CESM2",
    scenario = "ssp370",
    start_year = 2050,
    end_year = 2051,
    outdir = "met_drivers",
    resolution = "d01"
)

head(result)
```

By default, existing files are skipped. Set `overwrite = TRUE` only when files
should be regenerated.

```{r}
result <- cae_build_met_drivers(
    sites = sites,
    model = "CESM2",
    scenario = "ssp370",
    start_year = 2050,
    end_year = 2051,
    outdir = "met_drivers",
    resolution = "d01",
    overwrite = TRUE
)
```

## Output Structure

`cae_build_met_drivers()` writes files using this path pattern:

```text
{outdir}/{site_id}/{model}.{scenario}.{year}.nc
```

For example:

```text
met_drivers/fresno/CESM2.ssp370.2050.nc
met_drivers/los_angeles/CESM2.ssp370.2050.nc
met_drivers/sacramento/CESM2.ssp370.2050.nc
```

Each file contains one site and one year of hourly data. Dimensions are
`latitude`, `longitude`, and `time`. The time coordinate is stored as seconds
since the start of the file year.
