---
title: "Showcase"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Quick answer to "what can I actually do with this thing." Three questions you'd
really ask about California's climate future, each one knocked out in a few
lines of R straight off live Cal-Adapt data.

Plots aren't really the point. Point is, grabbing this data into R, which used
to be hours of grief, is already handled, so you get to spend your time on
science instead.

```{r}
#| message: false
library(caladaptaer)
```

## Comparing the two products

Analytics Engine ships two downscaled products, and they flat out don't line
up. WRF is dynamical, hourly, sits on a 45 km Lambert grid, and calls near
surface air temp `t2`. LOCA2 is statistical, daily, on a 3 km lat/lon grid, and
doesn't even carry a mean temperature, just daily `tasmax` and `tasmin`.
Different physics, different grid, different time step, even different variable
names.

caladaptaer hands you both back as one tidy `data.frame`, so you can just throw
them on the same plot. Here's the same GCM (EC-Earth3), SSP3-7.0, at Fresno, one
week of July 2050, pulled from both.

```{r}
# WRF: hourly t2, 45 km
wrf <- cae_fetch_point("t2", model = "EC-Earth3", scenario = "ssp370",
                       lon = -119.77, lat = 36.75,
                       start_time = "2050-07-01T00:00:00",
                       end_time   = "2050-07-07T23:00:00")

# LOCA2: daily tasmax and tasmin, 3 km, same model and spot
tmax <- cae_fetch_point("tasmax", model = "EC-Earth3", scenario = "ssp370",
                        lon = -119.77, lat = 36.75,
                        timescale = "day", resolution = "d03",
                        start_time = "2050-07-01", end_time = "2050-07-07")

tmin <- cae_fetch_point("tasmin", model = "EC-Earth3", scenario = "ssp370",
                        lon = -119.77, lat = 36.75,
                        timescale = "day", resolution = "d03",
                        start_time = "2050-07-01", end_time = "2050-07-07")

head(wrf, 3)
head(tmax, 3)
```

Same call both times. Past the variable name, only thing that changes is
`timescale = "day"` and `resolution = "d03"` to aim at LOCA2 store instead of
WRF one.

```{r}
#| fig-width: 8
#| fig-height: 4.5
#| fig-alt: "EC-Earth3 air temperature at Fresno for the first week of July 2050. A grey line traces hourly WRF 2 meter temperature swinging between roughly 11 and 39 C. Red and blue dashed lines mark LOCA2 daily tasmax near 37 to 40 C and tasmin near 20 to 24 C."
# kelvin to C
wrf$c  <- wrf$value  - 273.15
tmax$c <- tmax$value - 273.15
tmin$c <- tmin$value - 273.15

plot(wrf$time, wrf$c, type = "l", lwd = 2, col = "grey30",
     ylim = range(c(wrf$c, tmax$c, tmin$c)),
     xlab = "time", ylab = "temperature (C)")
lines(tmax$time, tmax$c, col = "#D62728", lty = 2); points(tmax$time, tmax$c, pch = 19, col = "#D62728")
lines(tmin$time, tmin$c, col = "#1F77B4", lty = 2); points(tmin$time, tmin$c, pch = 19, col = "#1F77B4")
legend("topright", bty = "n", cex = 0.8,
       legend = c("WRF t2 (hourly)", "LOCA2 tasmax (daily)", "LOCA2 tasmin (daily)"),
       col = c("grey30", "#D62728", "#1F77B4"),
       lwd = c(2, 1, 1), lty = c(1, 2, 2), pch = c(NA, 19, 19))
```

And yeah, they disagree, that's the honest read. WRF overnight lows run cooler
than LOCA2 `tasmin`. Part of it is that 45 km WRF cell smears over way more, and
way more varied, ground than that little 3 km LOCA2 cell, and part of it is that
dynamical and statistical downscaling are just different animals. Bottom line,
which product you grab changes your number. Pulling both into R side by side is
how you actually see that for yourself instead of taking somebody's word for it.

## How much does the model you pick matter

Eight GCMs get dynamically downscaled in WRF. They all run through the same
downscaling machinery, so stacking them up at one spot gives you a clean read on
structural uncertainty: how much of your answer is actual climate, versus just
which model spit it out.

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Hourly 2 meter air temperature at Fresno on 1 July 2050 from every WRF GCM under SSP3-7.0. The curves share a cool night, warm afternoon shape but fan out 10 C or more apart at the afternoon peak, with the hottest model near 44 C."
wrf_370 <- Filter(
  function(m) "ssp370" %in% cae_scenarios(activity = "WRF", model = m),
  setdiff(cae_models(activity = "WRF"), "ERA5")
)

series <- lapply(wrf_370, function(m) {
  d <- cae_fetch_point("t2", model = m, scenario = "ssp370",
                       lon = -119.77, lat = 36.75,
                       start_time = "2050-07-01T00:00:00",
                       end_time   = "2050-07-01T23:00:00")
  d$model <- m
  d$c <- d$value - 273.15
  d
})
series <- do.call(rbind, series)

cols <- hcl.colors(length(wrf_370), "Dark 3")
plot(series$time, series$c, type = "n", xlab = "time", ylab = "temperature (C)")
for (i in seq_along(wrf_370)) {
  d <- series[series$model == wrf_370[i], ]
  lines(d$time, d$c, col = cols[i], lwd = 2)
}
legend("topright", legend = wrf_370, col = cols, lwd = 2, cex = 0.7, bty = "n")
```

Models agree on the shape, cool night, hot afternoon, but they fan out hard on
the actual temperature, 10 C or more apart at the afternoon peak (the hottest,
MPI-ESM1-2-HR, tops 44 C while the coolest sits near 30). Clock is UTC, so that
midday dip is really the local overnight low. Anything you run downstream that
cares about temperature, a crop model, a fire index, a building load, an
ecosystem carbon flux, that spread is the error bar you're stuck with just from
picking a model. Pushing the whole ensemble through to a calibrated answer is on
our roadmap. Grabbing that ensemble in one place, today, is what makes it doable.

## What only hourly data shows

LOCA2 and the raw GCMs are daily at best. WRF is hourly, and that cracks open
something nothing else can: the shape of the day. Heat stress,
evapotranspiration, nighttime respiration, they all key off the sub daily
swing, not the daily mean. So here's that swing at three California landscapes
over a summer stretch in 2050.

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Hourly 2 meter air temperature for 1 to 3 July 2050 at three California sites under CESM2 SSP3-7.0. The coast stays in a narrow 15 to 26 C band, the Central Valley swings widely from about 12 to 31 C, and the Sierra site is coldest at roughly 4 to 17 C."
sites <- data.frame(
  name = c("Coast (San Francisco)", "Central Valley (Fresno)", "Sierra (Mammoth area)"),
  lon  = c(-122.42, -119.77, -119.03),
  lat  = c(  37.77,   36.75,   37.63),
  col  = c("#1F77B4", "#D62728", "#2CA02C")
)

series <- lapply(seq_len(nrow(sites)), function(i) {
  d <- cae_fetch_point("t2", model = "CESM2", scenario = "ssp370",
                       lon = sites$lon[i], lat = sites$lat[i],
                       start_time = "2050-07-01T00:00:00",
                       end_time   = "2050-07-03T23:00:00")
  d$c <- d$value - 273.15
  d$name <- sites$name[i]
  d
})
series <- do.call(rbind, series)

plot(series$time, series$c, type = "n", xlab = "time", ylab = "temperature (C)")
for (i in seq_len(nrow(sites))) {
  d <- series[series$name == sites$name[i], ]
  lines(d$time, d$c, col = sites$col[i], lwd = 2)
}
legend("topright", legend = sites$name, col = sites$col, lwd = 2, cex = 0.8, bty = "n")
```

Coast barely budges, ocean pins it in a tight band. Central Valley swings the
widest. Sierra site is coldest at every hour. That's exactly the physics you'd
expect, and it drops out of three `cae_fetch_point()` calls. No daily product
can show you that.

## Build a variable that isn't in the file

Cal-Adapt hands you wind as two components, eastward `u10` and northward `v10`,
not as a single speed. No problem, grab both and combine them. Here's the 10 m
wind at Fresno over a couple of July days.

```{r}
#| fig-width: 8
#| fig-height: 4.5
#| fig-alt: "Hourly 10 meter wind speed at Fresno for 1 to 2 July 2050 under CESM2 SSP3-7.0, computed from the u and v components, running between about 2 and 8.5 m/s with windier afternoons."
# eastward and northward components
u <- cae_fetch_point("u10", model = "CESM2", scenario = "ssp370", lon = -119.77, lat = 36.75,
                     start_time = "2050-07-01T00:00:00", end_time = "2050-07-02T23:00:00")
v <- cae_fetch_point("v10", model = "CESM2", scenario = "ssp370", lon = -119.77, lat = 36.75,
                     start_time = "2050-07-01T00:00:00", end_time = "2050-07-02T23:00:00")

# wind speed is just the magnitude
ws <- data.frame(time = u$time, speed = sqrt(u$value^2 + v$value^2))

plot(ws$time, ws$speed, type = "l", lwd = 2, col = "#2C7FB8",
     xlab = "time", ylab = "wind speed (m/s)")
```

Same trick works for anything you can build off the raw fields, vapor pressure
deficit, a heat index, reference ET. The parts are all in there.

## Zoom out to the whole state

Points are one thing, but you can read the whole grid and clip it to whatever
region you care about. Here's a summer afternoon over California: pull the 45 km
WRF field, clip it to the state, and you have a map and a number, the statewide
average, out of the same object.

```{r}
#| message: false
library(stars)
library(sf)
library(maps)

# one 45 km timestep over the western US
g <- cae_fetch("t2", model = "CESM2", scenario = "ssp370", resolution = "d01",
               start_time = "2050-07-01T00:00:00", end_time = "2050-07-01T00:00:00")

# California polygon, into the WRF grid's projection
ca <- st_transform(st_as_sf(map("state", "california", plot = FALSE, fill = TRUE)), st_crs(g))

# clip the grid to the state, kelvin to C
g_ca <- g[ca]
g_ca[[1]] <- units::drop_units(g_ca[[1]]) - 273.15

# statewide average out of one aggregate call
round(as.numeric(aggregate(g, ca, FUN = mean, na.rm = TRUE)[[1]])[1] - 273.15, 1)
```

```{r}
#| fig-width: 6
#| fig-height: 7
#| fig-alt: "Map of 2 meter air temperature clipped to California from the 45 km WRF grid for a summer 2050 afternoon under CESM2 SSP3-7.0. The Central Valley and southern deserts are hot, near 35 to 44 C, the coast and northern mountains cooler, near 15 to 25 C."
plot(g_ca, reset = FALSE, axes = TRUE,
     col = hcl.colors(100, "YlOrRd", rev = TRUE),
     key.pos = 4, key.width = lcm(1.4), key.length = 0.8, main = "")
plot(st_geometry(ca), add = TRUE, border = "black", lwd = 1.2)
```

That's 45 km, so California comes out to about 191 cells, blocky but honest. The
statewide mean falls out of one `aggregate()` call. Want a county or a watershed
instead? Swap the polygon, everything else stays the same.

## Where to go next

These analyses use only what the package does today. The
[home page](../index.html) has the roadmap, and
[Data Access](data_access.qmd) covers every function with live output.
