---
title: "Building SIPNET Met Drivers from WRF Projections"
---

```{r}
#| include: false
knitr::opts_chunk$set(eval = FALSE)
```

::: {.callout-note}
## What you will learn

- Which meteorological variables SIPNET requires and where they come from
  in the WRF dataset
- How WRF native units map to CF standard names and units
- How to download all 7 variables and write a CF-compliant NetCDF file
:::

## Background

[SIPNET](https://github.com/PecanProject/sipnet) is a process-based
ecosystem model that simulates carbon, water, and energy fluxes at hourly
timesteps. It expects met drivers in CF-standard NetCDF format.

The Cal-Adapt WRF hourly output is a natural fit for SIPNET met forcing
under future climate scenarios. This vignette walks through the pipeline:
query the catalog, read all variables, convert units, and write a NetCDF.

## Variable mapping

SIPNET needs 5 meteorological inputs plus wind components:

| CF standard name | WRF name | WRF units | CF units | Conversion |
|-----------------|----------|-----------|----------|------------|
| `air_temperature` | `t2` | K | K | none |
| `air_pressure` | `psfc` | Pa | Pa | none |
| `precipitation_flux` | `prec` | mm/timestep | kg/m2/s | divide by dt |
| `specific_humidity` | `q2` | kg/kg (mixing ratio) | kg/kg (specific) | q = w/(1+w) |
| `surface_downwelling_shortwave_flux_in_air` | `swdnb` | W/m2 | W/m2 | none |
| `eastward_wind` | `u10` | m/s | m/s | none |
| `northward_wind` | `v10` | m/s | m/s | none |

Most need no conversion. Precipitation goes from accumulated mm per timestep
to an instantaneous flux rate (dividing by 3600 for hourly data). Humidity
goes from mixing ratio to specific humidity -- the difference is small but
worth getting right.

## Download all variables

```{r}
library(caladaptR)

sipnet_vars <- c("t2", "psfc", "prec", "q2", "swdnb", "u10", "v10")

var_data <- lapply(sipnet_vars, function(v) {
  message("reading ", v)
  ca_fetch(v, model = "CESM2", scenario = "ssp370", n_timesteps = 24)
})
names(var_data) <- sipnet_vars
```

## Inspect the data

Always spot-check that values fall in physically reasonable ranges before
feeding them to an ecosystem model.

```{r}
# temperature: 200--350 K
summary(var_data$t2[[1]])

# pressure: 50000--110000 Pa
summary(var_data$psfc[[1]])

# shortwave: 0--1400 W/m2
summary(var_data$swdnb[[1]])
```

::: {.callout-important}
## Sanity checks

WRF output is model output, not observations. Temperature below 200 K,
negative shortwave, or negative pressure indicate a data problem --
don't pass those through to the model.
:::

## Write CF-standard NetCDF

`ca_write_cf_netcdf()` handles unit conversion and CF metadata in one call:

```{r}
ca_write_cf_netcdf(
  var_list = var_data,
  outfile = "CESM2.ssp370.2014-09-01.cf.nc",
  dt_seconds = 3600,
  overwrite = TRUE
)
```

The output includes CF-1.8 global attributes, standard variable names, and
computed wind speed. The output is ready for SIPNET ensemble runs.

## Point extraction alternative

For single-site runs, extract time series at a specific location instead of
writing the full grid:

```{r}
pts <- lapply(sipnet_vars, function(v) {
  ca_fetch_point(v, model = "CESM2", scenario = "ssp370",
                 lon = -119.77, lat = 36.75, n_timesteps = 24)
})
names(pts) <- sipnet_vars
```

![Hourly meteorological variables at Fresno, CA from the CESM2 SSP3-7.0
WRF simulation. These are the inputs SIPNET needs for a single-site
carbon flux simulation. The diurnal signatures are physically consistent
-- peak radiation drives peak temperature with a lag, humidity drops as
temperature rises, and pressure follows a semidiurnal
tide.](figures/fig05_met_panel_fresno.png){#fig-met width=95%}

## References

- Rahimi, S. et al. (2024). An overview of the Western United States
  Dynamically Downscaled Dataset (WUS-D3). *Geoscientific Model Development*,
  17, 2265--2286.
  [doi:10.5194/gmd-17-2265-2024](https://doi.org/10.5194/gmd-17-2265-2024)
- Dietze, M. C. et al. (2014). A quantitative assessment of a terrestrial
  biosphere model's data needs across North American biomes.
